Probabilistic Catalogs for Crowded Stellar Fields

### 2012-11-19 paper/ms.tex
More work on the intro, I feel it's pretty good now

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods such as maximum likelihood have also been used to model crowded fields
\citep[e.g.][]{irwin}. However, maximum likelihood is not completely appropriate
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise.

@@

To proceed with the model for inferring catalogs (stellar positions and fluxes)
from image data, we must
specify a definite hypothesis space and choices for the prior distribution
and the sampling distribution. These choices are presented and discussed in
Section~\ref{sec:model}. Throughout this paper we make a number of simplifications
for the purposes of testing the computational feasibility and performance of
the method. In real applications, some of these assumptions (such as known PSF
and noise variance) may be relaxed.

@@

\section{MCMC Implementation}
The MCMC sampling was implemented using the Diffusive Nested Sampling
\citep{dnest} method (hereafter DNS). DNS is a variant of the
Nested Sampling \citep{skilling} algorithm that uses Metropolis-Hastings
updates, and
is very generally applicable. The main difference between DNS and the standard
Metropolis-Hastings algorithm is that the target distribution is modified.
Rather than simply exploring the posterior distribution over catalog space,
DNS constructs an alternative target distribution which is a mixture of the
prior distribution with more constrained versions of the prior distribution.
The modified target distribution assists the sampling in several ways.
Firstly, the target distribution shrinks at a constant rate with time during the
initial phase of the exploration. This is similar to the popular ``simulated
annealing'' method but with an optimal annealing schedule. Secondly,
communication with the prior is maintained: once a catalog is found that
fits the data, the catalog can ``disintegrate'' back to the prior distribution
and re-fit, allowing different peaks in the parameter space to be explored
(if they exist). This all happens naturally within the context of a valid
MCMC sampler. The MCMC may also be run using the standard Metropolis algorithm
targeting the posterior distribution.

@@

\section{Comparison to SExtractor}
In the previous section we established that the inference of the catalogs from
the data is computationally feasible and that the number of stars and the
luminosity function can be inferred from the image data, albeit with moderate
uncertainty. We now compare this approach with the output from the standard
tool \sex. To achieve this, we executed \sex~on the two test cases, for various
values of the significance level cutoff. This results in a set of catalogs
for each image, with more conservative cutoffs resulting in less stars detected
as compared to more agressive cutoffs. The results for the number of stars in
the images are shown in Figure~\ref{fig:sex_N}. The main result is that
\sex~significantly underestimates the number of stars in the image for all
possible values of the cutoff parameter.

@@

\section{Discussion}

>>>>

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if an unknown-$N$ model had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods such as maximum likelihood have also been used to model crowded fields
\citep[e.g.][]{irwin}. However, maximum likelihood is not completely appropriate
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise.

@@

To proceed with the model for inferring catalogs (stellar positions and fluxes)
from image data, we must
specify a definite hypothesis space and choices for the prior distribution
and the sampling distribution. These choices are presented and discussed in
Section~\ref{sec:model}. In Section~\ref{sec:mcmc} we briefly discuss our
MCMC implementation. Section~\ref{sec:simulated_data} describes the tests we
carried out on simulated data, and a comparison with \sex~results is presented
in Section~\ref{sec:sex}. We conclude in Section~\ref{sec:conclusion}.
%Throughout this paper we make a number of simplifications
%for the purposes of testing the computational feasibility and performance of
%the method. In real applications, some of these assumptions (such as known PSF
%and noise variance) may be relaxed.

@@

\section{MCMC Implementation}\label{sec:mcmc}
The MCMC sampling was implemented using the Diffusive Nested Sampling
\citep{dnest} method (hereafter DNS). DNS is a variant of the
Nested Sampling \citep{skilling} algorithm that uses Metropolis-Hastings
updates, and
is very generally applicable. The main difference between DNS and the standard
Metropolis-Hastings algorithm is that the target distribution is modified.
Rather than simply exploring the posterior distribution over catalog space,
DNS constructs an alternative target distribution which is a mixture of the
prior distribution with more constrained versions of the prior distribution.
The modified target distribution assists the sampling in several ways.
Firstly, the target distribution shrinks at a constant rate with time during the
initial phase of the exploration. This is similar to the popular ``simulated
annealing'' method but with an optimal annealing schedule. Secondly,
communication with the prior is maintained: once a catalog is found that
fits the data, the catalog can ``disintegrate'' back to the prior distribution
and re-fit, allowing different peaks in the parameter space to be explored
(if they exist). This all happens naturally within the context of a valid
MCMC sampler. The MCMC may also be run using the standard Metropolis algorithm
targeting the posterior distribution.

@@

\section{Comparison to SExtractor}\label{sec:sex}
In the previous section we established that the inference of the catalogs from
the data is computationally feasible and that the number of stars and the
luminosity function can be inferred from the image data, albeit with moderate
uncertainty. We now compare this approach with the output from the standard
tool \sex. To achieve this, we executed \sex~on the two test cases, for various
values of the significance level cutoff. This results in a set of catalogs
for each image, with more conservative cutoffs resulting in less stars detected
as compared to more agressive cutoffs. The results for the number of stars in
the images are shown in Figure~\ref{fig:sex_N}. The main result is that
\sex~significantly underestimates the number of stars in the image for all
possible values of the cutoff parameter.

@@

\section{Discussion and Conclusions}\label{sec:conclusion}
